---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# ========================
# User-Defined Parameters
# ========================

```{r}
# Fold-change thresholds
fc_cutoff_primary <- 1.5 # primary analysis, |Log2FC| >= fc_cutoff
run_sensitivity <- TRUE
fc_cutoff_sensitivity <- 1 # sensitivity analysis, |Log2FC| >= fc_sens

# Input data file
data_file <- "HMVEC-D3  MDMAB231 and SKBR3 Artemisinine MS July 2024.xlsx"

# Protein Log2 fold-change column
protein_fc_col <- "scaled_prot"

# Protein-adjusted phosphosite Log2 fold-change column
phospho_fc_col <- "adj_scaled_norm_phos"

# Abundance columns for MA plots
# Protein abundance
protein_test_col <- "hmvec_d3_art_10" # Test sample
protein_control_col <- "hmvec_d3_control_10" # Control sample

# Phosphosite abundance
phospho_test_col <- "hmvec_d3_art_2" # Test sample
phospho_control_col <- "hmvec_d3_control_2" # Control sample

```

# =====================
# Pipeline Setup
# =====================

# Prepare environment and dataset
```{r}
# Load modules
library(tidyverse)
library(readxl)
library(here)
library(janitor)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pheatmap)
library(conflicted)

# Set conflict preferences
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)
conflicts_prefer(dplyr::distinct)
conflicts_prefer(dplyr::add_row)

# Load Data
raw_df <- read_excel(
  here("data", data_file),
  col_names = FALSE
)
```

# Clean & Filter Data (w/ column name QC)
```{r}
clean_df <- raw_df %>%
  row_to_names(row_number = 2) %>%
  clean_names() %>%
  dplyr::slice(-(1:3)) %>%
  type.convert(as.is = TRUE)

# Ensure parameter column names match cleaned dataset names
protein_fc_col <- make_clean_names(protein_fc_col)
phospho_fc_col <- make_clean_names(phospho_fc_col)
protein_test_col <- make_clean_names(protein_test_col)
protein_control_col <- make_clean_names(protein_control_col)
phospho_test_col <- make_clean_names(phospho_test_col)
phospho_control_col <- make_clean_names(phospho_control_col)

required_cols <- c(
  protein_fc_col = protein_fc_col,
  phospho_fc_col = phospho_fc_col,
  protein_test_col = protein_test_col,
  protein_control_col = protein_control_col,
  phospho_test_col = phospho_test_col,
  phospho_control_col = phospho_control_col
)

missing_cols <- names(required_cols)[
  !required_cols %in% names(clean_df)
]

if (length(missing_cols) > 0) {

  bad_values <- required_cols[missing_cols]

  msg <- paste(
    paste0(missing_cols, " -> '", bad_values, "'"),
    collapse = "\n"
  )

  stop(
    "Invalid column parameters:\n\n",
    msg,
    call. = FALSE
  )
}
```

# Define helper functions
```{r}
# 1. Core Flagging Function
flag_effect_size <- function(x, fc_cutoff) {
  !is.na(x) & abs(x) >= fc_cutoff
}
 
# 2. Protein Flagging Function
flag_proteins <- function(df, column, fc_cutoff, suffix = "") {

  if (!column %in% names(df)) {
    stop(paste("Column", column, "not found in dataframe"))
  }

  col_name <- paste0("prot_sig", suffix)
  df[[col_name]] <- flag_effect_size(df[[column]], fc_cutoff)

  df
}

# 3. Phosphosite Flagging Function
flag_phosphosites <- function(df, column, fc_cutoff, suffix = "") {

  if (!column %in% names(df)) {
    stop(paste("Column", column, "not found in dataframe"))
  }
  
  col_name <- paste0("phos_sig", suffix)
  df[[col_name]] <- flag_effect_size(df[[column]], fc_cutoff)

  df
}
```

# =======================
# Create analysis tables
# =======================

```{r}
# Select desired columns
analysis_df <- clean_df %>%
  select(
    accession = improved_accession,
    gene = gene_name,
    primary_protein_status = protein_pre_improvement_master_status,
    
    scaled_prot = all_of(protein_fc_col),
    adj_scaled_norm_phos = all_of(phospho_fc_col),
    
    prot_test = all_of(protein_test_col),
    prot_control = all_of(protein_control_col),
    
    phos_test = all_of(phospho_test_col),
    phos_control = all_of(phospho_control_col)
  )

# Filter for master protein entries for protein analysis
protein_df <- analysis_df %>%
  filter(primary_protein_status == "IsMasterProtein",
         !is.na(prot_test),
         !is.na(prot_control)) %>%
  distinct(accession, .keep_all = TRUE) %>%
  mutate(
    M_prot = scaled_prot,
    A_prot = (prot_test + prot_control) / 2
  ) %>%
  select(accession, gene, scaled_prot, M_prot, A_prot)

# Separate phosphosite df to parallel protein df
phospho_df <- analysis_df %>%
  filter(!is.na(phos_test),
         !is.na(phos_control),
         !is.na(accession)) %>%
  mutate(
    M_phos = adj_scaled_norm_phos,
    A_phos = (phos_test + phos_control) / 2
  ) %>%
  select(accession, gene, adj_scaled_norm_phos, M_phos, A_phos)
```

# ==================================
# Preliminary QC & Data Exploration
# ==================================

# QC Summary Table
```{r}
qc_summary <- tibble(
  metric = c(
    "Total rows (analysis table)",
    "Unique proteins",
    "Protein rows analysed",
    "Phosphosite rows analysed",
    "Protein missing (%)",
    "Phosphosite missing (%)",
    "Median protein log2FC",
    "Median phosphosite log2FC",
    paste0("Proteins exceeding primary FC cutoff (|log2FC| ≥ ", fc_cutoff_primary, ")"),
    paste0("Phosphosites exceeding primary FC cutoff (|log2FC| ≥ ", fc_cutoff_primary, ")")
  ),
  value = c(
    nrow(analysis_df),
    n_distinct(protein_df$accession),
    nrow(protein_df),
    nrow(phospho_df),
    round(mean(is.na(analysis_df$scaled_prot)) * 100, 1),
    round(mean(is.na(analysis_df$adj_scaled_norm_phos)) * 100, 1),
    round(median(protein_df$scaled_prot, na.rm = TRUE), 3),
    round(median(phospho_df$adj_scaled_norm_phos, na.rm = TRUE), 3),
    sum(abs(protein_df$scaled_prot) >= fc_cutoff_primary, na.rm = TRUE),
    sum(abs(phospho_df$adj_scaled_norm_phos) >= fc_cutoff_primary, na.rm = TRUE)
  )
)

if (run_sensitivity) {

  qc_summary <- qc_summary %>%
    dplyr::add_row(
      metric = paste0(
        "Proteins exceeding sensitivity FC cutoff (|log2FC| ≥ ",
        fc_cutoff_sensitivity,
        ")"
      ),
      value = sum(abs(protein_df$scaled_prot) >= fc_cutoff_sensitivity, na.rm = TRUE)
    ) %>%
    dplyr::add_row(
      metric = paste0(
        "Phosphosites exceeding sensitivity FC cutoff (|log2FC| ≥ ",
        fc_cutoff_sensitivity,
        ")"
      ),
      value = sum(abs(phospho_df$adj_scaled_norm_phos) >= fc_cutoff_sensitivity, na.rm = TRUE)
    )

}

qc_summary
```

# Protein Distribution
```{r}
ggplot(protein_df, aes(x = scaled_prot)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = c(-fc_cutoff_primary, fc_cutoff_primary),
             linetype = "dashed") +
  xlim(-5, 5) +
  theme_minimal() +
  labs(
    title = "Protein Log2 Fold Change Distribution",
    x = "Log2 Fold Change",
    y = "Count"
  )
```

# Phosphosite Distribution
```{r}
ggplot(phospho_df, aes(x = adj_scaled_norm_phos)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = c(-fc_cutoff_primary, fc_cutoff_primary),
             linetype = "dashed") +
  xlim(-5, 5) +
  theme_minimal() +
  labs(
    title = "Phosphosite Log2 Fold Change Distribution",
    x = "Log2 Fold Change",
    y = "Count"
  )
```

# ============================================
# Differential Expression (Fold-Change Based)
# ============================================

```{r}
# Primary Analysis
protein_df <- flag_proteins(
  protein_df,
  column = "scaled_prot",
  fc_cutoff = fc_cutoff_primary,
  suffix = "_primary"
)

protein_df <- protein_df %>%
  mutate(
    prot_category_primary = case_when(
      M_prot >= fc_cutoff_primary  ~ "Upregulated",
      M_prot <= -fc_cutoff_primary ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )

phospho_df <- flag_phosphosites(
  phospho_df,
  column = "adj_scaled_norm_phos",
  fc_cutoff = fc_cutoff_primary,
  suffix = "_primary"
)

phospho_df <- phospho_df %>%
  mutate(
    phos_category_primary = case_when(
      M_phos >= fc_cutoff_primary  ~ "Upregulated",
      M_phos <= -fc_cutoff_primary ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )

# Sensitivity Analysis
if (run_sensitivity) {

  protein_df <- flag_proteins(
    protein_df,
    column = "scaled_prot",
    fc_cutoff = fc_cutoff_sensitivity,
    suffix = "_sens"
  )

  phospho_df <- flag_phosphosites(
    phospho_df,
    column = "adj_scaled_norm_phos",
    fc_cutoff = fc_cutoff_sensitivity,
    suffix = "_sens"
  )
}

if (run_sensitivity) {

  protein_df <- protein_df %>%
    mutate(
      prot_category_sens = case_when(
        M_prot >= fc_cutoff_sensitivity  ~ "Upregulated",
        M_prot <= -fc_cutoff_sensitivity ~ "Downregulated",
        TRUE ~ "Not Significant"
      )
    )

  phospho_df <- phospho_df %>%
    mutate(
      phos_category_sens = case_when(
        M_phos >= fc_cutoff_sensitivity  ~ "Upregulated",
        M_phos <= -fc_cutoff_sensitivity ~ "Downregulated",
        TRUE ~ "Not Significant"
      )
    )

}
```

# ======================================
# MA Plot Generation - Primary Analysis
# ======================================

# Protein MA Plot
```{r}
ggplot(protein_df, aes(x = A_prot, y = M_prot)) +

  # background proteins
  geom_point(
    data = subset(protein_df, prot_category_primary == "Not Significant"),
    colour = "grey60",
    alpha = 0.4,
    size = 1.2
  ) +

  # flagged proteins
  geom_point(
    data = subset(protein_df, prot_category_primary != "Not Significant"),
    aes(color = prot_category_primary),
    size = 2
  ) +
  geom_hline(
    yintercept = c(-fc_cutoff_primary, fc_cutoff_primary),
    linetype = "dashed"
  ) +
  geom_smooth(method = "loess", se = FALSE, colour = "black", linewidth = 0.5) +
  scale_color_manual(values = c(
    "Upregulated" = "#D7191C",
    "Downregulated" = "#2C7BB6"
  )) +
  labs(
    title = "Protein MA Plot",
    x = "Mean Log2 Abundance (A)",
    y = "Log2 Fold Change (M)",
    color = "Category"
  ) +
  theme_minimal()
```

# Phosphosite MA Plot
```{r}
ggplot(phospho_df, aes(x = A_phos, y = M_phos)) +

  geom_point(
    data = subset(phospho_df, phos_category_primary == "Not Significant"),
    colour = "grey65",
    alpha = 0.5,
    size = 1.2
  ) +

  geom_point(
    data = subset(phospho_df, phos_category_primary != "Not Significant"),
    aes(color = phos_category_primary),
    size = 2
  ) +

  geom_hline(yintercept = 0, linewidth = 0.3) +

  geom_hline(
    yintercept = c(-fc_cutoff_primary, fc_cutoff_primary),
    linetype = "dashed"
  ) +

  geom_smooth(method = "loess", se = FALSE, colour = "black", linewidth = 0.5) +

  scale_color_manual(values = c(
    "Upregulated" = "#D7191C",
    "Downregulated" = "#2C7BB6"
  )) +

  labs(
    title = "Phosphosite MA Plot",
    x = "Mean Log2 Abundance (A)",
    y = "Log2 Fold Change (M)",
    color = "Category"
  ) +
  theme_minimal()
```

# ==========================================
# MA Plot Generation - Sensitivity Analysis
# ==========================================

# Protein MA Plot
```{r}
ggplot(protein_df, aes(x = A_prot, y = M_prot)) +

  # background proteins
  geom_point(
    data = subset(protein_df, prot_category_sens == "Not Significant"),
    colour = "grey60",
    alpha = 0.4,
    size = 1.2
  ) +

  # flagged proteins
  geom_point(
    data = subset(protein_df, prot_category_sens != "Not Significant"),
    aes(color = prot_category_sens),
    size = 2
  ) +
  geom_hline(
    yintercept = c(-fc_cutoff_sensitivity, fc_cutoff_sensitivity),
    linetype = "dashed"
  ) +
  geom_smooth(method = "loess", se = FALSE, colour = "black", linewidth = 0.5) +
  scale_color_manual(values = c(
    "Upregulated" = "#D7191C",
    "Downregulated" = "#2C7BB6"
  )) +
  labs(
    title = "Protein MA Plot",
    x = "Mean Log2 Abundance (A)",
    y = "Log2 Fold Change (M)",
    color = "Category"
  ) +
  theme_minimal()
```

# Phosphosite MA Plot
```{r}
ggplot(phospho_df, aes(x = A_phos, y = M_phos)) +

  geom_point(
    data = subset(phospho_df, phos_category_sens == "Not Significant"),
    colour = "grey65",
    alpha = 0.5,
    size = 1.2
  ) +

  geom_point(
    data = subset(phospho_df, phos_category_sens != "Not Significant"),
    aes(color = phos_category_sens),
    size = 2
  ) +

  geom_hline(yintercept = 0, linewidth = 0.3) +

  geom_hline(
    yintercept = c(-fc_cutoff_sensitivity, fc_cutoff_sensitivity),
    linetype = "dashed"
  ) +

  geom_smooth(method = "loess", se = FALSE, colour = "black", linewidth = 0.5) +

  scale_color_manual(values = c(
    "Upregulated" = "#D7191C",
    "Downregulated" = "#2C7BB6"
  )) +

  labs(
    title = "Phosphosite MA Plot",
    x = "Mean Log2 Abundance (A)",
    y = "Log2 Fold Change (M)",
    color = "Category"
  ) +
  theme_minimal()
```

# ====================================
# Ranked Hit Lists - Primary Analysis
# ====================================

# Proteins
```{r}
protein_hits_primary <- protein_df %>%
  filter(prot_category_primary != "Not Significant") %>%
  mutate(effect_size = abs(scaled_prot)) %>%
  arrange(desc(effect_size)) %>%
  mutate(rank = row_number()) %>%
  select(rank, accession, gene, scaled_prot, A_prot, effect_size, prot_category_primary)

protein_hits_primary
```

# Phosphosites
```{r}
phospho_hits_primary <- phospho_df %>%
  filter(phos_category_primary != "Not Significant") %>%
  mutate(effect_size = abs(adj_scaled_norm_phos)) %>%
  arrange(desc(effect_size)) %>%
  mutate(rank = row_number()) %>%
  select(rank, accession, gene, adj_scaled_norm_phos, A_phos, effect_size, phos_category_primary)

phospho_hits_primary
```

# ========================================
# Ranked Hit Lists - Sensitivity Analysis
# ========================================

# Proteins
```{r}
protein_hits_sens <- protein_df %>%
  filter(prot_category_sens != "Not Significant") %>%
  mutate(effect_size = abs(scaled_prot)) %>%
  arrange(desc(effect_size)) %>%
  mutate(rank = row_number()) %>%
  select(rank, accession, gene, scaled_prot, A_prot, effect_size, prot_category_sens)

protein_hits_primary
```

# Phosphosites
```{r}
phospho_hits_sens <- phospho_df %>%
  filter(phos_category_sens != "Not Significant") %>%
  mutate(effect_size = abs(adj_scaled_norm_phos)) %>%
  arrange(desc(effect_size)) %>%
  mutate(rank = row_number()) %>%
  select(rank, accession, gene, adj_scaled_norm_phos, A_phos, effect_size, phos_category_sens)

phospho_hits_sens
```